Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-25 appendix.qmd
Updated Figure 1 with reproduable codes

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(quollr)
library(tibble)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(patchwork)
library(colorspace)
library(kableExtra)
library(conflicted)
library(ggforce)

@@

```{r}
umap_scurve <- read_rds(file = here::here("data/s_curve/s_curve_umap.rds"))

scurve_scaled_obj <- gen_scaled_data(
  nldr_data = umap_scurve)

sc_ltr_pos <- c(0.08, 0.96)
sc_xlims <- c(-0.5, 1.47)
sc_ylims <- c(-0.32, 2.1)
```

<!--Full hexagon grid with UMAP data-->

```{r}
## Compute hexbin parameters
num_bins_x_scurve <- 7

## hexagon binning to have regular hexagons
hb_obj_scurve <- hex_binning(
  nldr_obj = scurve_scaled_obj, 
  bin1 = num_bins_x_scurve, 
  q = 0.1)

## Data set with all centroids
all_centroids_df <- hb_obj_scurve$centroids

## Generate all coordinates of hexagons
hex_grid <- hb_obj_scurve$hex_poly

a1_temp <- hb_obj_scurve$a1
a2_temp <- hb_obj_scurve$a2
l_temp <- quad(a=3, b = 2 * a2_temp, c = -(a2_temp^2 + a1_temp^2))

```{r}
# Code to draw illustration for notation
start_pt <- all_centroids_df |> 
  filter(hexID == 1)

d_rect <- tibble(x1min = 0, 
                 x1max = 1,
                 x2min = 0,
                 x2max = diff(scurve_scaled_obj$lim2)/diff(scurve_scaled_obj$lim1))

hex_grid_temp40 <- hex_grid |> 
  filter(hex_poly_id == 53)

c_start <- tibble(s1 = 0.03, s2 = -0.05)

a1 <- tibble(x = all_centroids_df$c_x[4],
             xend = all_centroids_df$c_x[5],
             y = all_centroids_df$c_y[29],
             yend = all_centroids_df$c_y[29],
             label = expression(a[1]))
a2 <- tibble(x = all_centroids_df$c_x[39],
             xend = all_centroids_df$c_x[39],
             y = all_centroids_df$c_y[39],
             yend = all_centroids_df$c_y[47],
             label = expression(a[2]))

l <- tibble(x = hex_grid_temp40$x[2],
            xend = hex_grid_temp40$x[3],
            y = hex_grid_temp40$y[2],
            yend = hex_grid_temp40$y[3],
            label = expression(l))

@@

                 color = "#bdbdbd") +
    geom_point(data = all_centroids_df, aes(

@@

      x = x1min  - c_start$s1, # 0 - 0.03, 
      y = -0.35, 
      xend = x1max - c_start$s1, #1 - 0.03, 
      yend = -0.35), #-0.35),

@@

                 color = "black")+
    annotate("text", x=0.5, y=-0.45, 
             label = expression(r[1]), color = "black", size = 8) +

@@

                 color = "black")+ 
    annotate("text", x=-0.35, y=1, 
             label = expression(r[2]), color = "black", size = 8) +

@@

             color = "black", hjust=-0.2, size = 8) +
    annotate("text", x=-0.18, y=-0.25, 

@@

```{r}
## Compute hexbin parameters
num_bins_x_scurve <- 3

## hexagon binning to have regular hexagons
hb_obj_scurve <- hex_binning(
  nldr_obj = scurve_scaled_obj, 
  bin1 = num_bins_x_scurve, 
  q = 0.1)

## Data set with all centroids
all_centroids_df <- hb_obj_scurve$centroids

## Generate all coordinates of hexagons
hex_grid <- hb_obj_scurve$hex_poly

## Data to generate circle
circles <- data.frame(
  x = c(0.726, 0.451),
  y = c(1.27, 0.790),
  r = rep(0.32, 2)
)

emb_df1 <- tibble(emb1 = 0.61, emb2 = 0.99) |>
  add_row(emb1 = 0.54, emb2 = 1.08)

assign_data_vis1 <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "#bdbdbd", aes(group = hex_poly_id)) +
  geom_point(data = emb_df1, aes(x = emb1, y = emb2), color = "black", size = 0.7) +
  geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID), size = 5, colour = "#bdbdbd") +
  geom_circle(data = circles, aes(x0 = x, y0 = y, r = r), fill = NA) +
  geom_segment(x = 0.61, y = 0.99, xend = 0.46, yend = 0.83, 
               arrow = arrow(length = unit(0.02, "npc")), color = "#969696") +
  geom_segment(x = 0.54, y = 1.08, xend = 0.7, yend = 1.2, 
               arrow = arrow(length = unit(0.02, "npc")), color = "#969696")

hex_data <- hex_grid |> 
  dplyr::filter(hex_poly_id == 8)

@@

assign_data_vis2 <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "#bdbdbd", aes(group = hex_poly_id)) +
  geom_point(data = points_df, aes(x = x, y = y), color = "black", size = 0.7) +
  geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID), size = 5, colour = "#bdbdbd") +
  geom_segment(x = 0.175, y = 0.790, xend = -0.06, yend = 0.790, 
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") +
  geom_segment(x = 0.726, y = 0.790, xend = 0.50, yend = 0.790,
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") +
  geom_segment(x = 0.313, y = 1.03, xend = 0.41, yend = 0.82,
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") +
  geom_segment(x = 0.588, y = 1.03, xend = 0.48, yend = 0.82,
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") +
  geom_segment(x = 0.313, y = 0.551, xend = 0.20, yend = 0.34,
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") +
  geom_segment(x = 0.588, y = 0.551, xend = 0.67, yend = 0.34,
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") +
  geom_segment(x = 0.451, y = 1.11, xend = 0.451, yend = 0.85,
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") +
  geom_segment(x = 0.175, y = 0.949, xend = -0.02, yend = 0.82,
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") +
  geom_segment(x = 0.175, y = 0.631, xend = 0.175, yend = 0.4,
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") +
  geom_segment(x = 0.451, y = 0.472, xend = 0.25, yend = 0.3,
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") +
  geom_segment(x = 0.726, y = 0.631, xend = 0.726, yend = 0.4,
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") +
  geom_segment(x = 0.726, y = 0.949, xend = 0.5, yend = 0.82,
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd")

```{r}
#| label: fig-assign-data
#| fig-cap: "Binning the data. Points are assigned to the nearest centroid. If a point is equidistant from multiple centroids, assigned to the centroid with the smallest bin ID."
#| fig-width: 9
#| fig-height: 6
#| fig-pos: H

@@

error_umap_plot_pbmc <-  plot_mse(error_pbmc) +
  scale_x_continuous(breaks = sort(unique(error_pbmc$a1))[c(1, 5, 9, 12, 15, 17, 19, 20)]) +
  scale_color_manual(values=c('#377eb8', '#e41a1c'))
```

>>>>

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(tibble)
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(patchwork)
library(colorspace)
library(kableExtra)
library(conflicted)
library(ggforce)

@@

```{r}
#| label: code-illustration
# Code to draw illustration for notation
## hexagon binning to have regular hexagons

hex_grid_temp <- read_rds("data/two_nonlinear/hex_grid_temp.rds")
all_centroids_df_temp <- read_rds("data/two_nonlinear/all_centroids_df_tempp.rds")
start_pt <- read_rds("data/two_nonlinear/start_pt.rds")
d_rect <- read_rds("data/two_nonlinear/d_rect.rds")
a1 <- read_rds("data/two_nonlinear/a2_data.rds")
a2 <- read_rds("data/two_nonlinear/a1_data.rds")
l <- read_rds("data/two_nonlinear/l_data.rds")
rect_adj <- read_rds("data/two_nonlinear/rect_adj.rds")

@@

                 colour = "#bdbdbd") +
    geom_point(data = all_centroids_df_temp, aes(

@@

      x = x1min  - rect_adj$x1, # 0 - 0.03, 
      y = -0.31, 
      xend = x1max - rect_adj$x1, #1 - 0.03, 
      yend = -0.31), #-0.35),

@@

                 colour = "black")+
    annotate("text", x=0.5, y=-0.36, 
             label = expression(r[1]), colour = "black", size = 8) +

@@

                 colour = "black")+ 
    annotate("text", x=-0.3, y=0.4, 
             label = expression(r[2]), colour = "black", size = 8) +

@@

             colour = "black", hjust=-0.2, size = 8) +
    annotate("text", x=-0.18, y=-0.24, 

@@

```{r}

## Data set with all centroids
all_centroids_df <- read_rds("data/two_nonlinear/two_nonlinear_all_centroids_two_curvy4.rds")

## Generate all coordinates of hexagons
hex_grid <- read_rds("data/two_nonlinear/two_nonlinear_hex_grid_two_curvy4.rds")

## Data to generate circle
circles <- data.frame(
  x = c(0.608, 0.372),
  y = c(0.306, 0.715),
  r = rep(0.27, 2)
)

emb_df1 <- tibble(emb1 = 0.45, emb2 = 0.47) |>
  add_row(emb1 = 0.54, emb2 = 0.56)

assign_data_vis1 <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "#bdbdbd", aes(group = hex_poly_id)) +
  geom_point(data = emb_df1, aes(x = emb1, y = emb2), color = "black", size = 0.7) +
  geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID), size = 5, colour = "#bdbdbd") +
  geom_circle(data = circles, aes(x0 = x, y0 = y, r = r), fill = NA) +
  geom_segment(x = 0.45, y = 0.47, xend = 0.58, yend = 0.3, 
               arrow = arrow(length = unit(0.02, "npc")), color = "#969696") +
  geom_segment(x = 0.54, y = 0.56, xend = 0.4, yend = 0.715, 
               arrow = arrow(length = unit(0.02, "npc")), color = "#969696")
```

```{r}
hex_data <- hex_grid |> 
  dplyr::filter(hex_poly_id == 8)

@@

arrow_df <- tibble(
  x = points_df$x, 
  y = points_df$y,
  xend = c(0.37, points_df$x[2] - 0.1, 0.136, points_df$x[4] - 0.1, 0.61, points_df$x[6] - 0.1, 0.33, points_df$x[8] - 0.1, 0,  points_df$x[10] - 0.1, 0.58, 0.45),
  yend = c(0.82, points_df$y[2] - 0.1, 0.4, points_df$y[4] - 0.1, 0.4, points_df$y[6] - 0.1, 0.82, points_df$y[8] - 0.1, 0.715, points_df$y[10] - 0.1, 0.4, 0.71)
)

assign_data_vis2 <- ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "#bdbdbd", aes(group = hex_poly_id)) +
  geom_point(data = points_df, aes(x = x, y = y), color = "black", size = 0.7) +
  geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID), size = 5, colour = "#bdbdbd") +
  geom_segment(data = arrow_df, aes(x = x, y = y, xend = xend, yend = yend), 
               arrow = arrow(length = unit(0.02, "npc")), color = "#bdbdbd") 
```

```{r}
#| label: fig-assign-data
#| fig-cap: "Binning the data. Points are assigned to the nearest centroid. If a point is equidistant from multiple centroids, assigned to the centroid with the smallest bin ID."
#| fig-width: 12
#| fig-height: 6
#| fig-pos: H

@@

error_umap_plot_pbmc <-  plot_rmse(error_pbmc) +
  scale_x_continuous(breaks = sort(unique(error_pbmc$a1))[c(1, 5, 9, 12, 15, 17, 19, 20)]) +
  scale_color_manual(values=c('#377eb8', '#e41a1c'))
```

### 2025-06-25 appendix.tex
Updated Figure 1 with reproduable codes

\begin{tabular}{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
\textbf{Notation} & \textbf{Description}\\
\midrule
$n, p, k$ & number of observations, variables, embedding dimension, respectively\\
$\mathbfit{X}, \mathbfit{x}$ & $p$-dimensional data (population, sample)\\
$\mathbfit{y}$ & $k$-dimensional layout\\
$P$ & orthonormal basis, generating a $d\text{-}dimensional$ linear projection of $p$-dimensional data\\
$T$ & true  model\\
\addlinespace
$g$ & functional mapping from \pD{} to \kD{}, especially as prescribed by NLDR\\
$\mathbfit{\theta}$ & (Hyper-) parameters for NLDR method\\
$r$ & ranges of the embedding components\\
$C^{(j)}$ & $j$-dimensional bin centers\\
$(b_1, b_2)$ & number of bins in each direction\\
\addlinespace
$(a_1, a_2)$ & binwidths, distance between centroids in each direction\\
$(s_1, \ s_2)$ & starting coordinates of the hexagonal grid\\
$q$ & buffer to ensure hexgrid covers data, proportion of data range, 0-1\\
$m$ & number of non-empty bins\\
$b$ & number of  hexagons in the grid\\
\addlinespace
$h$ & hexagonal id\\
$l$ & side length\\
$A$ & area\\
\bottomrule
\end{tabular}
\endgroup{}

@@

@@

\includegraphics[width=0.3\linewidth,height=\textheight,keepaspectratio]{appendix_files/figure-pdf/fig-hex-param-1.pdf}

\caption{\label{fig-hex-param}The components of the hexagon grid
illustrating notation.}

@@

\caption{\label{fig-assign-data}Binning the data. Points are assigned to
the nearest centroid. If a point is equidistant from multiple centroids,
assigned to the lowest centroid.}

@@

Following the process in \citet{xia2023} again using the Human
Peripheral Blood Mononuclear Cells (PBMC) data. Note that
\citet{xia2023} uses a different subset from the PBMC dataset used by
\citet{chen2024}, which contains \(31,021\) cells including cell type
labels, and the gene expression levels were in the unit of
log-transformed UMI count per \(10,000\). They focused on three
sequencing methods (inDrops, DropSeq, and SeqWell) and four common cell
types Cytotoxic T cell, CD4+T cell, CD14+ Monocyte, and B cell.

>>>>

\begin{tabular}{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
\textbf{Notation} & \textbf{Description}\\
\midrule
$n, p, k$ & number of observations, variables, embedding dimension, respectively\\
$\mathbfit{X}, \mathbfit{x}$ & $p$-dimensional data (population, sample)\\
$\mathbfit{y}$ & $k$-dimensional layout\\
$P$ & orthonormal basis, generating a $d\text{-}dimensional$ linear projection of $p$-dimensional data\\
$T$ & true  model\\
$g$ & functional mapping from \pD{} to \kD{}, especially as prescribed by NLDR\\
$\mathbfit{\theta}$ & (Hyper-) parameters for NLDR method\\
$r$ & ranges of the embedding components\\
$C^{(j)}$ & $j$-dimensional bin centers\\
$(b_1, b_2)$ & number of bins in each direction\\
$(a_1, a_2)$ & binwidths, distance between centroids in each direction\\
$(s_1, \ s_2)$ & starting coordinates of the hexagonal grid\\
$q$ & buffer to ensure hexgrid covers data, proportion of data range, 0-1\\
$m$ & number of non-empty bins\\
$b$ & number of  hexagons in the grid\\
$h$ & hexagonal id\\
$l$ & side length\\
$A$ & area\\
$n_h$ & bin count\\
$w_h$ & standardised bin counts\\
\bottomrule
\end{tabular}
\endgroup{}

@@

\section{Generating the nonlinear clusters
data}\label{generating-the-nonlinear-clusters-data}

@@

@@

\caption{\label{fig-assign-data}Binning the data. Points are assigned to
the nearest centroid. If a point is equidistant from multiple centroids,
assigned to the centroid with the smallest bin ID.}

@@

Note that \citet{xia2023} uses a different subset from the PBMC dataset
used by \citet{chen2024}, shown by us in the main paper example, which
is why this comparison is shown here and not in the main paper. Their
data contains \(31,021\) cells including cell type labels, and the gene
expression levels were in the unit of log-transformed UMI count per
\(10,000\). They focused on three sequencing methods (inDrops, DropSeq,
and SeqWell) and four common cell types Cytotoxic T cell, CD4+T cell,
CD14+ Monocyte, and B cell. Pre-processing follows the process in
\citet{xia2023} again using the Human Peripheral Blood Mononuclear Cells
(PBMC) data.

### 2025-06-25 paper.qmd
Updated Figure 1 with reproduable codes

<!--Fit the best model for author suggestion and compute error-->
<!--pbmc3k/14_gen_model_with_UMAP.R-->
```{r}
#| label: hexbin-pbmc
#| 
umap_pbmc_scaled_with_cluster <- read_rds("data/pbmc3k/umap_pbmc_scaled_with_cluster.rds")

@@

<!--pbmc3k/11_gen_mse_for_diff_methods.R-->
```{r}
#| label: combine-data-pbmc

@@

<!--best choice-->
<!--Fit the best model and compute error-->
<!--pbmc3k/15_gen_model_with_tSNE.R-->
```{r}
#| label: tsne-num-bins-pbmc
tsne_pbmc_scaled_best_with_cluster <- read_rds("data/pbmc3k/tsne_pbmc_scaled_best_with_cluster.rds")

>>>>

<!--Fit the best model for author suggestion and compute error-->
<!--pbmc3k/13_gen_model_with_UMAP.R-->
```{r}
#| label: hexbin-pbmc
#| 
umap_pbmc_scaled_with_cluster <- read_rds("data/pbmc3k/umap_pbmc_scaled_with_cluster.rds")

@@

<!--pbmc3k/08_gen_mse_for_diff_methods.R-->
```{r}
#| label: combine-data-pbmc

@@

<!--best choice-->
<!--Fit the best model and compute error-->
<!--pbmc3k/14_gen_model_with_tSNE.R-->
```{r}
#| label: tsne-num-bins-pbmc
tsne_pbmc_scaled_best_with_cluster <- read_rds("data/pbmc3k/tsne_pbmc_scaled_best_with_cluster.rds")
